---
title: 'Heart Attack Dataset: An Analysis'
author: "Ananthajit Srikanth"
date: "19/07/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Heart attacks are the leading cause of death worldwide (source: Our World in Data). Hence, understanding how to predict the possibility of heart disease is extremely important. 

This analysis will use the UCI Machine Learning Repository's Heart Disease Dataset to predict the risk of heart disease in patients.

## Dataset Information

This analysis uses four datasets, each of which are from four different regions:

* Hungarian Institute of Cardiology
* University Hospital, Zurich, Switzerland and University Hospital, Basel, Switzerland
* Cleveland Clinic Foundation
* V.A. Medical Center, Long Beach

Each of these regions form the *processed.hungarian*,*processed.switzerland*,*processed.cleveland*, and *processed.va* datasets.

The *processed* prefix indicates that out of 76 variables, only 14 are used. All the tables are also comma separated.

Given below are the variables in the dataset, as per the heart-disease.names file:

1. Age in years
2. Sex (assigned at birth) (1 is male, and 0 is female)
3. Chest Pain Type (one of four values):
  1 - Typical Angina
  2 - Atypical Angina
  3 - Non anginal Pain
  4 - Asymptomatic
4. trestbps - Resting Blood Pressure in mm Hg
5. chol - Serum Cholesterol in mg/dl
6. fbs - Is fasting blood sugar greater than 120 mg/dl? (1 = yes, 2 = no)
7. restecg - Resting Electrocardiogram (ECG) results (one of three values):
  0 - Normal
  1 - Having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV): This is a symptom of Myocardial Infarction
  2 - Showing probable or definite left ventricular hypertrophy by Estes' criteria (enlargment of the left ventricle, usually to compensate for tissue loss)
8. thalach - Maximum Heart Rate Achieved (in beats per minute)
9. exang - Exercise Induced Angina (1 = yes, 0 = no)
10. oldpeak - ST depression induced by exercise relative to rest
11. slope - Slope of ST segment (as seen in ECG) during excercise:
  * 1 - Upsloping
  * 2 - Flat
  * 3 - Downsloping
12. ca - Number of vessels colored by fluoroscopy (0 - 3) 
13. thal: One of three possible values:
  * 3 - Normal
  * 6 - Fixed Defect
  * 7 - Reversible Defect

14. num (outcome) - Diagnosis of Heart Disease:
  1 - More than 50% narrowing of diameter of any major heart vessel
  0 - Less than 50% narrowing of diameter of any major heart vessel.  
  
## Creating the Datasets

The datasets are publicly available at the [UCI Machine Learning repository website](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/) 

First, libraries have to be loaded for analysis:

```{r}
# Heart attack analysis

# Loading the required libraries, can add required libraries later in vector (same code as used in previous project, source: Ananthajit Srikanth)
required_libraries <- c("caret", "matrixStats", "tidyverse", "knitr", "lubridate", "broom", "randomForest", "rattle", "rpart")
# The for loop installs and loads libraries one by one from required_libraries
for(i in 1:length(required_libraries)){
  if(!require(required_libraries[i], character.only = TRUE)){
    install.packages(required_libraries[i], repos = "http://cran.us.r-project.org")
    library(required_libraries[i], character.only = TRUE)
  }
  else{
    require(required_libraries[i], character.only = TRUE)}}

# Run this line only if necessary:
# update.packages()
```

Once the libraries are loaded, the datasets can be downloaded:

```{r}


```




# Bibliography
1. Hungarian Institute of Cardiology. Budapest: Andras Janosi, M.D.
       2. University Hospital, Zurich, Switzerland: William Steinbrunn, M.D.
       3. University Hospital, Basel, Switzerland: Matthias Pfisterer, M.D.
       4. V.A. Medical Center, Long Beach and Cleveland Clinic Foundation:
	  Robert Detrano, M.D., Ph.D.
https://ourworldindata.org/causes-of-death
https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/