---
title: 'Heart Attack Dataset: An Analysis'
author: "Ananthajit Srikanth"
date: "19/07/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Heart attacks are the leading cause of death worldwide (source: Our World in Data). Hence, understanding how to predict the possibility of heart disease is extremely important.

This analysis will use the UCI Machine Learning Repository's Heart Disease Dataset to predict the risk of heart disease in patients.

## Dataset Information

This analysis uses four datasets, each of which are from four different regions:

* Hungarian Institute of Cardiology
* University Hospital, Zurich, Switzerland and University Hospital, Basel, Switzerland
* Cleveland Clinic Foundation
* V.A. Medical Center, Long Beach

Each of these regions form the *processed.hungarian*,*processed.switzerland*,*processed.cleveland*, and *processed.va* datasets.

The *processed* prefix indicates that out of 76 variables, only 14 are used. All the tables are also comma separated.

Given below are the variables in the dataset, as per the heart-disease.names file:

1. Age in years
2. Sex (assigned at birth) (1 is male, and 0 is female)
3. Chest Pain Type:
  1 - Typical Angina
  2 - Atypical Angina
  3 - Non anginal Pain
  4 - Asymptomatic
4. trestbps - Resting Blood Pressure in mm Hg
5. chol - Serum Cholesterol in mg/dl
6. fbs - Is fasting blood sugar greater than 120 mg/dl? (1 = yes, 2 = no)
7. restecg - Resting Electrocardiogram (ECG) results:
  0 - Normal
  1 - 
8.
9.
10.
11.
12.
13.
14.
    
  





```{r}

```




# Bibliography
1. Hungarian Institute of Cardiology. Budapest: Andras Janosi, M.D.
       2. University Hospital, Zurich, Switzerland: William Steinbrunn, M.D.
       3. University Hospital, Basel, Switzerland: Matthias Pfisterer, M.D.
       4. V.A. Medical Center, Long Beach and Cleveland Clinic Foundation:
	  Robert Detrano, M.D., Ph.D.
https://ourworldindata.org/causes-of-death